---
title: "Site C Point Count Data: Songbird Diversity Analyses"
output:
  html_notebook:
    toc: TRUE
    toc_float: TRUE
---

Exploration of bird community metrics using Site C bird point count survey data.

What do I want to do?\
-Species richness and diversity\
-Expected diversity using rarefaction\
-Community structure exploration using ordination and cluster analysis\
-Indicator species analysis

# Setup

```{r message=FALSE}
# Packages
library(tidyverse)
library(vegan)
library(BiodiversityR)
library(vcd)
library(indicspecies)         # Indicator species analysis
library(lme4)
library(glmmTMB)
library(MuMIn)
library(emmeans)
library(effects)
library(DHARMa)

```

Import the tidied raw songbird station and observation data, if no already in memory.

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
options(scipen = 999)
```

# Approach

Need to determine how repeated samples will be dealt with.

For patterns of species richness, want to use sample (survey) and use site (station) as random effect. That is clear.

But what about other community analysis about composition, like sp accumulation and ordination? Could roll up surveys to station. The problem with this approach is that there are differnt number of surveys per station and any roll-up would introduce an effort bias. I think samples (surveys) must be used for all analyses.

Questions:

# Import and prep

Load the 5-min count data. Note for later: May need to create a surveyID (single variable identifier).

```{r Import}
song_wide_5min <- read_csv(here("point_counts", "data_processed", "song_wide_5min.csv"))

# Survey covariats
survey_cov <- read_csv(here("point_counts", "data_processed", "survey_cov.csv")) %>% 
  mutate(Year = as_factor(Year))
  
```

Create the sp matrices for use in vegan.

```{r}
sp <- song_wide_5min %>% 
  select(`B-ALFL`:`B-YRWA`)
```

Prepare env data and create lumped BHC variables.

```{r}
env <- song_wide_5min %>% 
  select(`StationID`:`SurveyDuration`) %>%
  mutate(Year = as_factor(Year)) %>% 
  mutate(BHC17 = case_when(
    BHC20 == "FBT" ~ "FBS/FBT",
    BHC20 == "FBS" ~ "FBS/FBT",
    BHC20 == "DSG" ~ "DSG/DSS",
    BHC20 == "DSS" ~ "DSG/DSS",
    TRUE ~ BHC20
    )) %>% 
  mutate(BHC17 = factor(BHC17, levels = c("CSH","CYF","CMF","DSH","DYF","DMF",
                                          "RSH","RYF","RMF", "FBS/FBT","WGR",
                                          "WSH","WRI","DSG/DSS","CUL","NVE", 
                                          "ANT")
                        )
         ) %>% 
  mutate(BHC8 = case_when(
    BHC20 == "CSH"	~ "Conif",
    BHC20 == "CYF"	~ "Conif",
    BHC20 == "CMF"	~ "Conif",
    BHC20 == "DSH"	~ "Decid",
    BHC20 == "DYF"	~ "Decid",
    BHC20 == "DMF"	~ "Decid",
    BHC20 == "RSH"	~ "RipForest",
    BHC20 == "RYF"	~ "RipForest",
    BHC20 == "RMF"	~ "RipForest",
    BHC20 == "FBS"	~ "Wetland",
    BHC20 == "FBT"	~ "Conif",
    BHC20 == "WGR"	~ "Wetland",
    BHC20 == "WSH"	~ "Wetland",
    BHC20 == "WRI"	~ "Wetland",
    BHC20 == "DSG"	~ "Dry slopes",
    BHC20 == "DSS"	~ "Dry slopes",
    BHC20 == "CUL"	~ "Cultiv",
    BHC20 == "NVE"	~ "Non-veg",
    BHC20 == "ANT"	~ "Anthropogenic")
    ) %>% 
  mutate(BHC8 = factor(BHC8, levels = c("Conif", "Decid", 'RipForest', 
                                            "Wetland", "Dry slopes", "Cultiv", 
                                            "Non-veg", "Anthropogenic")
                       )
         ) %>% 
  left_join(survey_cov, by = c("StationID", "Visit", "Year", "Date", "Time", 
                               "SurveyDuration"))
```

# Patterns of species richness

```{r paged.print=TRUE}
rich <- diversityresult(sp, index = "richness", method = "each site", sortit = FALSE, digits = 6)
shan <- diversityresult(sp, index = "Shannon", method = "each site", sortit = FALSE, digits = 6)
simp <- diversityresult(sp, index = "Simpson", method = "each site", sortit = FALSE, digits = 6)
divdat <- bind_cols(env, rich, shan, simp)

hist(divdat$richness)
hist(divdat$Shannon)
hist(divdat$Simpson)

# or
ggplot(divdat, aes(x=richness)) + 
  geom_histogram(binwidth = 1)

```

The richness data seems to fit poisson a bit bettter.

```{r paged.print=TRUE}

fit_poisson <- goodfit(divdat$richness, type = "poisson")
rootogram(fit_poisson)
summary(fit_poisson)

fit_nbinom <- goodfit(divdat$richness, type = "nbinomial")
rootogram(fit_nbinom)
summary(fit_nbinom)

```

Not sure how to make sense of the goodness of fit tests on Shannon div.

```{r paged.print=TRUE}

fit_poisson <- goodfit(divdat$Shannon, type = "poisson")
rootogram(fit_poisson)
summary(fit_poisson)

fit_nbinom <- goodfit(divdat$Shannon, type = "nbinomial")
rootogram(fit_nbinom)
summary(fit_nbinom)

```

## Model

```{r paged.print=TRUE}
model_full <- glmmTMB(richness ~ Year + BHC8 + EastWest + TSSR + JDAY + 
                        (1|StationID),
                   data = divdat,
                   family = (poisson("log"))
                   )

# Not sur eif this is needed
# na.action = "na.fail")

dredge(model_full)

```

For the model with lowest AICc
```{r}

model_sel <- glmmTMB(richness ~ Year + BHC8 + EastWest + JDAY + 
                        (1|StationID),
                   data = divdat,
                   family = (poisson("log"))
                   )

summary(model_sel)

ae <- allEffects(model_sel)
plot(ae, residuals="TRUE")

```




## A test with Year as continuous
```{r paged.print=TRUE}
model_sel <- glmmTMB(richness ~ MonYear + BHC8 + EastWest + JDAY + 
                        (1|StationID),
                   data = divdat,
                   family = (poisson("log"))
                   )

summary(model_sel)

ae <- allEffects(model_sel)
plot(ae, residuals="TRUE")
```

# Species accumulation / rarefaction

iNext is the package for this. There is some ore though,

Basic, using vegan.

```{r}
sac <- specaccum(sp)
plot(sac, ci.type="polygon", ci.col="yellow")

```

# Ordination

```{r}

```

# Indicator Species Analysis

```{r}

```
